Nondeterministic Bigraphical Reactive Systems for Markov Decision Processes\thanks{Supported by organization x.}

### 2018-08-15 paper.tex
An OK-ish draft

% TODO: opening statement on why probabilities and non-determinism are important
\begin{abstract}
  In this paper we introduce two extensions of bigraphical reactive systems
  (BRSs) to handle probabilistic and nondeterministic behaviour, allowing BRSs
  to be used in discrete-time Markov chain and Markov decision process (MDP)
  generation. We extend the implementation of an open-source tool for BRSs
  BigraphER with new syntax for MDP actions and rewards, better visualisation
  capabilities, and data export options. Furthermore, we introduce a new
  front-end for working with BRSs based on Jupyter notebooks, allowing the user
  to quickly visualise bigraphs and reaction rules, simulate execution, or
  produce a complete transition system, all within the notebook interface.
  Finally, we demonstrate the modelling capabilities of the new system with a
  case study on the movement and behaviour of autonomous agents, exploring ideas
  such as collecting objects, tracking visited locations, hierarchies of spaces,
  and uncertainty in the topology of space.

@@

% TODO: introduction
%Bigraphical reactive systems (BRS) \cite{DBLP:books/daglib/0022395} are a
%universal formalism for modelling interacting systems that evolve in time and
%space.

@@

% TODO xavier \cite{koenig1998xavier}
\begin{definition}[Markov decision process] \label{mdp}
  For any finite set $X$, let $\Dist(X)$ denote the set of discrete probability
  distributions over $X$. A \emph{Markov Decision Process} is a tuple $ (S,
  \overline{s}, A, P, L)$, where: $S$ is a finite set of states and
  $\overline{s} \in S$ is the initial state; $A$ is a finite set of
  \emph{actions}; $P : S \times A \to \Dist(S)$ is a (partial) probabilistic
  transition function, mapping state-action pairs to probability distributions
  over $S$; $L : S \to 2^P$ is a labelling with atomic propositions.
\end{definition}

@@

We begin by introducing some notation used in the definitions. Natural numbers
are sometimes interpreted as finite ordinals, i.e., $m = \{ 0, \dots, m - 1 \}$.
We write $S \uplus T$ to indicate the union of sets known or assumed to be
disjoint. For two functions $f$ and $g$ with disjoint domains $S$ and $T$ we
write $f \uplus g$ for the function with domain $S \uplus T$ such that $(f
\uplus g)|_S = f$ and $(f \uplus g)|_T = g$. The identity function on the set
$S$ is written as $\Id_S$. Given a binary relation $\mathit{rel} \subseteq A
\times B$, we denote the domain restriction of $\mathit{rel}$ over $S \subseteq
A$ by $S \lhd \mathit{rel}$. Similarly, we write $\mathit{rel} \rhd S$ for the
range restriction of $\mathit{rel}$ over $S \subseteq B$. We also denote the
transitive closure of $\mathit{rel}$ by $\mathit{rel}^+$. In defining bigraphs
we assume that names, node identifiers, edge identifiers, and controls are drawn
from four infinite pairwise disjoint sets, respectively $\mathcal{X},
\mathcal{V}, \mathcal{E}$, and $\mathcal{K}$.

@@

Informally, a bigraph can be thought of as a graph, where a node can be inside
another node\footnote{To be specific, in this paper we extend the ideas behind
  \emph{bigraphs with sharing}
  \cite{DBLP:phd/ethos/Sevegnani12,dblp:journals/tcs/sevegnanic15}, where a node
  can be contained in the intersection of several other nodes.}. However, the
pictures tend to become much clearer if we instead visualise the nesting
relations between nodes as just a different type of edges. Consider the bigraph
in Fig.~\ref{example_bigraph}. The white ellipses are the \emph{nodes}. Each
node has a type, called \emph{control}, and denoted here by the labels
$\mathsf{A}, \mathsf{B}$, and $\mathsf{C}$. The green edges are actually
hyperedges (can connect any positive number of nodes), and are called
\emph{links}. In this example they are given names $\mathsf{x}, \mathsf{y}$, and
$\mathsf{e}$. Each node has an ordered list of \emph{ports} (not pictured in the
diagram) that can be thought of as sockets for links. Nodes of the same control
have the same number of ports. The number of ports a node of control
$\mathsf{K}$ can have is called the \emph{arity} of $\mathsf{K}$ and is denoted
by $\ar(\mathsf{K})$. The two white dashed rectangles represent \emph{regions},
and in this representation are the parents to any otherwise parentless nodes.
Grey dashed rectangles are called \emph{sites}, and encode parts of the model
that have been abstracted away. Nodes, sites, and regions are also known as
\emph{places}. The directed black arrows between places show the containment
relations, e.g., the $\mathsf{A}$ node has two children (inner nodes):
$\mathsf{B}$ and the only site in this bigraph, labelled by the number $0$.

@@

% TODO: mention we will be using bigraphs with sharing and we recall some
% definitions from TCS paper
\begin{definition}[concrete place graph with sharing]
  A \emph{concrete place graph with sharing}
  \[ F = (V_F, \ctrl_F, \prnt_F) : m \to n \]
  is a triple having an inner interface $m$ and an outer interface $n$. These
  index the sites and regions of the place graph, respectively. $F$ has a finite
  set $V_F \subset \mathcal{V}$ of nodes, a control map $\ctrl_F : V_F \to
  \mathcal{K}$, and a \emph{parent relation}
  \[ \prnt_F \subseteq (m \uplus V_F) \times (V_F \uplus n) \]
  that is acyclic, i.e., $(v, v) \not\in \prnt_F^+$ for any $v \in V_F$.
\end{definition}

@@

% TODO: for the same D_i - not clear
% TODO: minimal example showing normalisation
\begin{definition}[probabilistic reaction rule]
  A \emph{probabilistic reaction rule} $\mathsf{R}$ is a triple $(R, R', p)$,
  sometimes written $R \xrightarrowRHD{p} R'$, where $(R, R')$ is a reaction
  rule and $p \in (0, 1]$ is a probability. Similarly to
  Definition~\ref{reaction_rule}, it generates a set of ground reaction rules of
  the form $(r, r', p)$.
\end{definition}

@@

  Let $g, g'$ be ground bigraphs, and $\{ (r_i, r'_i, p_i) \}_{i=1}^n$ a set of
  ground probabilistic reaction rules, where for each $r_i$, there exists a
  bigraph $D_i$ such that $g = D_i \circ r_i$. Let $S = \{ (r_i, r'_i, p_i) \mid
  g' = D_i \circ r'_i \}$ (for the same $D_i$), and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation is defined as
  \[ g \xrightarrowrhd{p}_{\mathcal{R}} g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', p') \in S} p'. \]
\end{definition}

@@

% TODO: intro explaining how we are further generalising the extensions defined
% above

\begin{definition}[nondeterministic reaction rule]
  Let $A$ be a set of actions. A \emph{nondeterministic reaction rule}
  $\mathsf{R}$ is a tuple $(R, R', a, p)$, where $(R, R', p)$ is a probabilistic
  reaction rule, and $a \in A$ is an action. We also define a \emph{reaction
    reward function} $r : A \to \mathbb{R}$ that assigns a reward or cost to
  each action.
\end{definition}

% TODO: D_i not clear
% TODO: different symbol for reward
\begin{definition}[nondeterministic bigraphical reactive system (NBRS)]
  A \emph{nondeterministic bigraphical reactive system} consists of a pair
  $(\mathcal{B}, \mathcal{R})$, where $\mathcal{B}$ is a set of ground bigraphs
  and $\mathcal{R}$ is a set of nondeterministic reaction rules defined over
  $\mathcal{B}$.

  Let $g, g'$ be ground bigraphs, $a \in A$ an action, and $\{ (r_i, r'_i, a,
  p_i) \}_{i=1}^n$ a set of ground nondeterministic reaction rules with action
  $a$, where for each $r_i$, there exists a bigraph $D_i$ such that $g =
  D_i \circ r_i$. Let $S = \{ (r_i, r'_i, a, p_i) \mid g' = D_i \circ r'_i \}$
  (for the same $D_i$), and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation for action $a$ is defined as
  \[ g \xrightarrowrhd[r(a)]{p}_{a} g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', a, p') \in S} p'. \]
\end{definition}

@@

% TODO: minimal example, possible the same one as below

@@

% TODO: we extend BigraphER
% TODO: describe the new syntactical constructs using the example
% TODO: what's the reward of action b?

@@

The MDP can be represented as an NBRS with BigraphER code in
Listing~\ref{example_listing}. More specifically, reaction rule probabilities
are represented as floating-point numbers inside the arrows (e.g.
\texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in squared brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\};}).

@@

\subsection{Front end}

@@

In order to demonstrate the use of bigraphs in modelling probabilistic
processes with nondeterminism, we present two examples of agent movement
planning in two dimensions. The examples have different sources of probabilistic
behaviour and explore topics such as tracking visited locations and uncertainty
about the topology.

% TODO: the agent is simply modelled as an atomic node as the focus of this case
% study is on its movements across a grid. More complex models are possible
% by...

@@

% TODO: taking advantage of the hierarchical structure of bigraphs

Our second model explores two important concepts in modelling the topology of
space: nesting and uncertainty. Nesting is the idea that spaces can be inside
other spaces: rooms inside floors, floors inside buildings, buildings inside
streets, etc. A key characteristic of nested spaces is that there is a separate
action (or several) taking you from the outer space to the inner space or vice
versa (e.g., using a door to enter or leave a room). In the rest of the paper,
whenever we discuss one space nested inside another, we will refer to the inner
space as the \emph{room}. While nesting has been modelled using bigraphs before
\cite{DBLP:conf/giscience/WaltonW12}, our approach models each room not as a
single entity, but rather as an interconnected collection of smaller spaces,
which we continue to call \emph{cells}. Thus, it is not enough to be in the
hallway in order to enter a room, one needs to be next to the door. Similarly,
after entering a room, the agent is positioned next to the other side of the
door.

@@

% TODO: limitation of a single entrance/exit per cell, can a room have multiple
% doors (doesn't make sense for this conceptualisation of the problem)?

% TODO: state diagrams (and statistics about them)
% TODO: numerical analysis (or ideas)
% TODO: discussion section
% TODO: what are the adtantages of using NBRS?
% TODO: limitations
% TODO: difficulties

@@

% TODO advantages: more visual/realistic than a vector of numbers, rewards and
% transition rules can be generated with a higher degree of automation, less
% code, more flexible (changing size of the grid); disadvantages: constant
% probability (without hacking)
% Extensions: multiple agents (nice), uncertainty in actions (easy), rotations
% (pomdp paper, easy), not limited to four directions

>>>>

\begin{abstract}
  Probabilities and nondeterminism allow us to represent uncertainty and
  decision making. In this paper we introduce two extensions of bigraphical
  reactive systems (BRSs) to handle probabilistic and nondeterministic
  behaviour, allowing BRSs to be used in discrete-time Markov chain and Markov
  decision process (MDP) generation. We extend the implementation of an
  open-source tool for BRSs BigraphER with new syntax for MDP actions and
  rewards, better visualisation capabilities, and data exporting options.
  Furthermore, we introduce a new front end for working with BRSs based on
  Jupyter notebooks, allowing the user to quickly visualise bigraphs and
  reaction rules, simulate execution, or produce a complete transition
  system---all within the notebook interface. Finally, we demonstrate the
  modelling capabilities of the new system with a case study on the movement and
  behaviour of autonomous agents, exploring ideas such as collecting objects,
  tracking visited locations, hierarchies of spaces, and uncertainty in the
  topology of space.

@@

TODO: introduction to bigraphs and everything else
%Bigraphical reactive systems (BRS) \cite{DBLP:books/daglib/0022395} are a
%universal formalism for modelling interacting systems that evolve in time and
%space.

@@

\begin{definition}[Markov decision process] \label{mdp}
  For any finite set $X$, let $\Dist(X)$ denote the set of discrete probability
  distributions over $X$. A \emph{Markov Decision Process} is a tuple $(S,
  \overline{s}, A, P, L)$, where: $S$ is a finite set of states and
  $\overline{s} \in S$ is the initial state; $A$ is a finite set of
  \emph{actions}; $P : S \times A \to \Dist(S)$ is a (partial) probabilistic
  transition function, mapping state-action pairs to probability distributions
  over $S$; $L : S \to 2^P$ is a labelling with atomic propositions.
\end{definition}

@@

In this paper we use \emph{bigraphs with sharing} and recall some of the
definitions from the literature
\cite{DBLP:phd/ethos/Sevegnani12,dblp:journals/tcs/sevegnanic15}. We begin by
introducing the notation used in the definitions. Natural numbers are sometimes
interpreted as finite ordinals, i.e., $m = \{ 0, \dots, m - 1 \}$. We write $S
\uplus T$ to indicate the union of sets known or assumed to be disjoint. For two
functions $f$ and $g$ with disjoint domains $S$ and $T$ we write $f \uplus g$
for the function with domain $S \uplus T$ such that $(f \uplus g)|_S = f$ and
$(f \uplus g)|_T = g$. The identity function on the set $S$ is written as
$\Id_S$. Given a binary relation $\mathit{rel} \subseteq A \times B$, we denote
the domain restriction of $\mathit{rel}$ over $S \subseteq A$ by $S \lhd
\mathit{rel}$. Similarly, we write $\mathit{rel} \rhd S$ for the range
restriction of $\mathit{rel}$ over $S \subseteq B$. We also denote the
transitive closure of $\mathit{rel}$ by $\mathit{rel}^+$. In defining bigraphs
we assume that names, node identifiers, edge identifiers, and controls are drawn
from four infinite pairwise disjoint sets, respectively $\mathcal{X},
\mathcal{V}, \mathcal{E}$, and $\mathcal{K}$.

@@

Informally, a bigraph can be thought of as a graph, where a node can be inside
another node (or in the intersection of several nodes). However, the diagrams
tend to become much clearer if we instead visualise the nesting relations
between nodes as just a different type of edges. Consider the bigraph in
Fig.~\ref{example_bigraph}. The white ellipses are the \emph{nodes}. Each node
has a type, called \emph{control} and denoted here by the labels $\mathsf{A},
\mathsf{B}$, and $\mathsf{C}$. The green edges are actually hyperedges (can
connect any positive number of nodes) and are called \emph{links}. In this
example they are given names $\mathsf{x}, \mathsf{y}$, and $\mathsf{e}$. Each
node has an ordered list of \emph{ports} (not pictured in the diagram) that can
be thought of as sockets for links. Nodes of the same control have the same
number of ports. The number of ports a node of control $\mathsf{K}$ can have is
called the \emph{arity} of $\mathsf{K}$ and is denoted by $\ar(\mathsf{K})$. The
two white dashed rectangles represent \emph{regions} and in this representation
are the parents to any otherwise parentless nodes. Grey dashed rectangles are
called \emph{sites} and encode parts of the model that have been abstracted
away. Nodes, sites, and regions are also known as \emph{places}. The directed
black arrows between places show the containment relations, e.g., the
$\mathsf{A}$ node has two children (inner nodes): $\mathsf{B}$ and the only site
in this bigraph, labelled by the number $0$.

@@

\begin{definition}[concrete place graph with sharing]
  A \emph{concrete place graph with sharing}
  \[ F = (V_F, \ctrl_F, \prnt_F) : m \to n \]
  is a triple having an inner interface $m$ and an outer interface $n$. These
  index the sites and regions of the place graph, respectively. $F$ has a finite
  set $V_F \subset \mathcal{V}$ of nodes, a control map $\ctrl_F : V_F \to
  \mathcal{K}$, and a \emph{parent relation}
  \[ \prnt_F \subseteq (m \uplus V_F) \times (V_F \uplus n) \]
  that is acyclic, i.e., $(v, v) \not\in \prnt_F^+$ for any $v \in V_F$.
\end{definition}

@@

\begin{definition}[probabilistic reaction rule]
  A \emph{probabilistic reaction rule} $\mathsf{R}$ is a triple $(R, R', p)$,
  sometimes written $R \xrightarrowRHD{p} R'$, where $(R, R')$ is a reaction
  rule and $p \in (0, 1]$ is a probability. Similarly to
  Definition~\ref{reaction_rule}, it generates a set of ground reaction rules of
  the form $(r, r', p)$.
\end{definition}

@@

  Let $g, g'$ be ground bigraphs, and $\{ (r_i, r'_i, p_i) \}_{i=1}^n$ a set of
  ground probabilistic reaction rules, where for each $r_i$, there exists a
  bigraph $D(r_i)$ such that $g = D(r_i) \circ r_i$. Let $S = \{ (r_i, r'_i, p_i) \mid g' = D(r_i) \circ r'_i \}$, and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation is defined as
  \[ g \xrightarrowrhd{p}_{\mathcal{R}} g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', p') \in S} p'. \]
\end{definition}

TODO: minimal example showing normalisation

@@

%The next step towards...
% TODO: intro explaining how we are further generalising the extensions defined
% above

\begin{definition}[nondeterministic reaction rule]
  Let $A$ be a set of actions. A \emph{nondeterministic reaction rule}
  $\mathsf{R}$ is a tuple $(R, R', a, p)$, where $(R, R', p)$ is a probabilistic
  reaction rule, and $a \in A$ is an action. We also define a \emph{reaction
    reward function} $\rho : A \to \mathbb{R}$ that assigns a reward or cost to
  each action.
\end{definition}

\begin{definition}[nondeterministic bigraphical reactive system (NBRS)]
  A \emph{nondeterministic bigraphical reactive system} consists of a pair
  $(\mathcal{B}, \mathcal{R})$, where $\mathcal{B}$ is a set of ground bigraphs
  and $\mathcal{R}$ is a set of nondeterministic reaction rules defined over
  $\mathcal{B}$.

  Let $g, g'$ be ground bigraphs, $a \in A$ an action, and $\{ (r_i, r'_i, a,
  p_i) \}_{i=1}^n$ a set of ground nondeterministic reaction rules with action
  $a$, where for each $r_i$, there exists a bigraph $D(r_i)$ such that $g =
  D(r_i) \circ r_i$. Let $S = \{ (r_i, r'_i, a, p_i) \mid g' = D(r_i) \circ r'_i
  \}$, and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation for action $a$ is defined as
  \[ g \xrightarrowrhd[\rho(a)]{p}_{a} g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', a, p') \in S} p'. \]
\end{definition}

@@

TODO: minimal example, possibly the same one as below.

@@

@@

The MDP can be represented as an NBRS with BigraphER code in
Listing~\ref{example_listing}. More specifically, reaction rule probabilities
are represented as floating-point numbers inside the arrows (e.g.
\texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in square brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\};}).

@@

\subsection{Front End}

@@

In order to demonstrate the use of bigraphs in modelling probabilistic
processes with nondeterminism, we present two examples of agent movement
planning in two dimensions. The examples have different sources of probabilistic
behaviour and explore topics such as tracking visited locations and uncertainty
about the topology. In both cases the agent is simply modelled as an single
node, as the focus of this case study is on its movement. More complex models
are possible by considering which direction the agent is facing
\cite{koenig1998xavier} or adding extra state to the agent such as limited
battery life or high/lower power sensors \cite{dblp:conf/nfm/giaquintahimn18}.

@@

Our second model explores two important concepts in modelling the topology of
space: nesting and uncertainty. Nesting is the idea that spaces can be inside
other spaces: rooms inside floors, floors inside buildings, buildings inside
streets, etc. A key characteristic of nested spaces is that there is a separate
action (or several) taking you from the outer space to the inner space or vice
versa (e.g., using a door to enter or leave a room). In the rest of the paper,
whenever we discuss one space nested inside another, we will refer to the inner
space as the \emph{room}. As bigraphs possess their own node nesting, this makes
bigraphical representations of these spaces more intuitive and easier to manage
than, say, a vector of numbers, each of which is assigned a particular meaning.
While nesting has been modelled using bigraphs before
\cite{DBLP:conf/giscience/WaltonW12}, our approach models each room not as a
single entity, but rather as an interconnected collection of smaller spaces,
which we continue to call \emph{cells}. Thus, it is not enough to be in the
hallway in order to enter a room, one needs to be next to the door. Similarly,
after entering a room, the agent is positioned next to the other side of the
door.

@@

\section{Discussion}

TODO (possibly): state transition diagrams and statistics about the number of
states, etc.
TODO: quantitative analysis with PRISM (need good nontrivial examples)

\begin{itemize}
\item LIMITATIONS
  \begin{itemize}
  \item no easy way to change probabilities depending on the state
  \item each room only has a single entrance/exit (could be fixed via
    bigraphs with sharing)
  \item can't implement everything you can think of: no way to check for set
    membership, limited support for representing numbers (adding/subtracting $n$
    has to be a separate reaction rule for each $n$)
  \end{itemize}
\item EXTENSIONS
  \begin{itemize}
  \item discovery of spaces independent of room structure (no specific ideas
    yet, but could be very nice)
  \item there's no good reason why rewards are assigned to actions rather than
    reaction rules
  \item multiple agents
  \item uncertainty in actions (like in Artificial Intelligence: A Modern
    Approach)
  \end{itemize}
\item ADVANTAGES
  \begin{itemize}
  \item less code than with PRISM
  \item more visual and intuitive, represents the modelled situation in a more
    straightforward way
  \item rewards and transition rules can be generated with a higher degree of
    automation (considering a specific property of a state, rather than listing
    all states that satisfy that property)
  \item more flexible in changing the size of the grid
  \item can be easily extended to more than four directions
  \end{itemize}
\end{itemize}

@@
